A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-11 tex/codatadp.tex
experimental stuff

@@

{\footnotesize
\begin{verbatim}
For the evaluation of Section 5, I'm playing with various ideas. I certainly
would like to keep it short and not waste too much of our time on it. We
discussed earlier having (1) hand-crafted problems and (2) actual benchmarks
(like those you already have). I suggest we drop (1); I never did hand-crafted
before and this always looks suspicious anyway. For (2), I see three or four
possible benchmark sets:

1. without (co)datatype theories
2. with datatypes
(3. with codatatypes)
4. with both

The benchmark problems would be identical except for the declarations at the
top ("declare-sort" and "declare-fun" vs. "declare-(co)datatype").
Irrespective of whether interpreted (co)datatypes are used, Sledgehammer will
typically include some (co)datatype axioms (i.e. tautologies if (co)datatypes
are enabled). Hence the tests are a good indication of whether native
(co)datatypes are useful to Sledgehammer. I assume the results will follow
common sense, i.e. 4 > {3, 2} > 1

As for the actual Isabelle theory files to use, for codatatypes there is not
much question: We just use all I could get my hands on. But perhaps we should
have a second benchmark set, ideally something a bit standard like, say, the
Judgment Day (JD) suite. Or I can make some arbitrary a priori (before
evaluation) decisions, but that always looks a bit fishy unless it can be
justified easily. The trouble is, we have about 1000 datatypes lying around in
Isabelle and the Archive of Formal Proofs (AFP). These would have no
codatatypes in them, so we could only do 1 and 2 of the evaluation. We could
at least try the "list" library, which is extremely developed and would have
1600 goals already, and perhaps a few really basic types ("option", "string"),
or perhaps red-black trees (about 450 goals).

So we can show the success rates, in a table like this:

                 Codatatype-oriented suite      Datatype (or JD) suite
                      (2345 goals)                   (1240 goals)

   w/o (co)data        1255                            699
   w data              1268                            712
   w codata            1270
   w both              1273

I can write the usual warning that an increase of a couple of percentage
points in success rate in a general tool like Sledgehammer and general
benchmark suites means a lot etc. Putting the datatype evaluations (i.e. 2 and
4) in there offers the advantage that it will give some measure of how much to
expect from the codatatypes. I.e. if the success rate doesn't go up by 50% for
datatypes, nobody will expect this to happen for codatatypes. A priori, I
would expect a similar gain, but I'm looking forward to see the experimental
data.

In addition to success rate (or number of solved goals), we should have other
statistics, like how often do we get proofs that contain the cyclicity and
bisimilarity rules. And finally, we look in more detail at one or two such
proofs, by presenting it and explaining it (and, before that, understanding
it).
\end{verbatim}}

>>>>

  * how many problems
  * how many problems have some codatatypes, how many have datatypes (including
    XXX that have both)

@@

  * wanted to evaluate both the impact of datatypes and codatatypes
  * datatypes are known to be very useful on SMT-LIB, e.g. blah
  * but we wanted to see how much they make a difference in Isabelle

  * ran all problems four times, to try all four combinations of datatypes and
    codatatypes enabled or disabled

  * lemmas/axioms often include basic properties of constructor and selectors,
    which are tautologies and ignored if (co)datatypes are enabled
  * allows a fair comparison: how much do (co)datatype support brings as
    opposed to just letting Sledgehammer do what it would normally do

  * experience with such features is not extremely high -- e.g. arithmetic,
    a most useful theory on lots of benchmarks, increases Sledgehammer's
    success rate by 2 percentage points, or 4\%, with Z3 in earlier work \cite{xxx}

  * so when analyzing statistics of this kind, with highly varied problems,
    we must not overestimate the importance of a single trick
  * on the other hand: every percentage point counts! cite Tom Hales

{\footnotesize
\begin{verbatim}
So we can show the success rates, in a table like this:

                 Datatype suite     Mixed suite        Codatatype suite
                (2345 problems)    (XXX problems)       (1240 problems)

   w/o (co)data        800           1255                699
   w data              815           1268                
   w codata                          1270                712
   w both                            1273
\end{verbatim}}

* statistics, like how often do we get proofs that contain the cyclicity and
  bisimilarity rules

* and finally, we look in more detail at one or two such proofs, by presenting
  it and explaining it (and, before that, understanding it)
